---
title: 'Avoid Brittle Tests: Best Practices with shinytest2 for Shiny Apps'
description: 'Learn how to avoid brittle tests by not testing implementation details with shinytest2 in your Shiny applications.'
pubDate:  'May 30 2024'
tags: ["r", "shiny", "shinytest2", "tests"]
---

`shinytest2` is a powerful tool, but with great power comes great responsibility.

`shinytest2` allows you to easily test a Shiny app, which is a web app, without having to switch to JavaScript – the language of the web.

However, the methods that `shinytest2::AppDriver` implements, are closely coupled to the inner workings of Shiny. Methods like:
- `set_inputs`
- `get_values`
- `click`
they all allow you to tap directly into the values of the Shiny server.

It can be great for apps that are already built and lack tests. With `shinytest2::record_test` you can rapidly add a lot of testing scenarios.

But when you're developing an app, and you're expecting it to change, tying tests so closly to implementation details will produce tests that you'll need to refactor often.

```{r}
test_app <- function() {
  get_data <- \(name) {
    list(
      iris = iris,
      mtcars = mtcars
    ) |>
      purrr::pluck(name, .default = data.frame())
  }
  shiny::shinyApp(
    ui = shiny::fluidPage(
      shiny::selectInput("dataset", "Dataset", c("Iris" = "iris", "Mtcars" = "mtcars")),
      shiny::tableOutput("preview")
    ),
    server = function(input, output, session) {
      output$preview <- shiny::renderTable({
        head(get_data(input$dataset))
      })
    }
  )
}
```

There is an app that displays a preview of a dataset based on the selection. The options in the dropdown have been renamed, so that users see capitalized names of datasets.

Let's see how we can test it with the `shinytest2::AppDriver`.

```{r}
library(testthat)

open_page <- function(app = test_app()) {
  page <- shinytest2::AppDriver$new(app)
  withr::defer(page$stop(), parent.frame(2))
  page
}

describe("DatasetSelect", {
  it("should show preview of the selected dataset", {
    # Arrange
    page <- open_page()

    # Act
    page$set_inputs(dataset = "mtcars")

    # Assert
    s <- selenider::selenider_session(driver = page, local = FALSE)
    s |>
      selenider::find_element("#preview > table > thead > tr") |>
      selenider::elem_children() |>
      testthat::expect_length(11)
  })
})
```

Let's disregard the difficult to understand assertion and focus on the problems with **Act**.

This test doesn't interact with the system from the perspective of a user. A user sees "Mtcars" not "mtcars". We need to know the underlying value to be able to set the input. This is what testing using implementation details means.

To make the test more robust, we should not rely on knowing the inner workings of the code we're testing.

Let's see how we can make it better.

The first thing would be to find how to select a value in the dropdown, but using values that a user sees.

```{r}
DatasetPreview <- R6::R6Class(
  public = list(
    driver = NULL,
    selenider = NULL,
    initialize = function(app) {
      self$driver <- shinytest2::AppDriver$new(app)
      self$selenider <- selenider::selenider_session(
        driver = self$driver,
        local = FALSE
      )
    },
    select = function(name) {
      self$selenider |>
        selenider::find_element("#dataset + div") |>
        selenider::elem_click()
      self$selenider |>
        selenider::find_elements("#dataset + div .option") |>
        selenider::elem_find(\(x) selenider::has_text(x, name)) |>
        selenider::elem_click()
    },
    expect_summary_n_cols = function(n) {
      self$driver$wait_for_idle()
      self$selenider |>
        selenider::find_element("#preview > table > thead > tr") |>
        selenider::elem_children() |>
        testthat::expect_length(n)
    },
    close = function() {
      self$driver$stop()
    }
  )
)

open_page <- function(app = test_app()) {
  page <- DatasetPreview$new(app)
  # withr::defer(page$close(), parent.frame(2))
  page
}

describe("DatasetSelect", {
  it("should show preview of the selected dataset", {
    # Arrange
    page <- open_page()

    # Act
    page$select("Mtcars")

    # Assert
    page$expect_summary_n_cols(11)
  })
})
```

Let's see what happened.



Other topic to cover:

What stubbing in tests might tell you - dependency injection.
